---
title: "Flight Delays, Weather Events, and Public Sentiment Analysis"
authors: "Yasmine Graham, Derek Fernandez-Davies, Abidur Rahman, Mojo Lindor, Colton Metcalf"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)



```

# 1. Libraries

```{r libraries}
# Text mining and sentiment analysis (Chapter 14 and 15)
library(tidytext)
# Data manipulation and visualization (Chapter 12) 
library(tidyverse)
# Data reshaping
library(tidyr)
# Date handling
library(lubridate)
# Nice tables
library(knitr)
```

# 2. Data Loading

```{r data_loading}
# Load all three main datasets

# This is the Flights dataset 2015 flight delays from Kaggle
flights <- read.csv("data/flights.csv")

# Supporting lookup tables for flights
airlines <- read.csv("data/airlines.csv")
airports <- read.csv("data/airports.csv")

# This is the Weather dataset NOAA 2015 storm events 
storm_data <- read.csv("data/storm_events_cleaned.csv")

# This is the Tweets dataset airline sentiment tweets 
twitter_data <- read.csv("data/Tweets-1.csv")

#check of all dataset dimensions
cat("Flights rows:", nrow(flights), "\n")
cat("Storm events rows:", nrow(storm_data), "\n")
cat("Tweets rows:", nrow(twitter_data), "\n")
cat("Airlines rows:", nrow(airlines), "\n")
cat("Airports rows:", nrow(airports), "\n")
```

# 3. Data Cleaning

## Flights

```{r flights_cleaning}
#Full airline names by joining with airlines lookup table
flights_clean <- flights %>%
  left_join(airlines, by = c("AIRLINE" = "IATA_CODE")) %>%
  rename(AIRLINE_NAME = AIRLINE.y) %>%
  # State info by joining airports on origin airport code
  left_join(airports %>% select(IATA_CODE, STATE),
            by = c("ORIGIN_AIRPORT" = "IATA_CODE")) %>%
  rename(ORIGIN_STATE = STATE) %>%
  # Removed rows with no delay info
  filter(!is.na(DEPARTURE_DELAY)) %>%
  # Replaced NA delay values with 0
  # NA in delay columns means no delay occurred
  mutate(
    DEPARTURE_DELAY = ifelse(is.na(DEPARTURE_DELAY), 0, DEPARTURE_DELAY),
    ARRIVAL_DELAY = ifelse(is.na(ARRIVAL_DELAY), 0, ARRIVAL_DELAY),
    WEATHER_DELAY = ifelse(is.na(WEATHER_DELAY), 0, WEATHER_DELAY),
    AIR_SYSTEM_DELAY = ifelse(is.na(AIR_SYSTEM_DELAY), 0, AIR_SYSTEM_DELAY),
    AIRLINE_DELAY = ifelse(is.na(AIRLINE_DELAY), 0, AIRLINE_DELAY),
    LATE_AIRCRAFT_DELAY = ifelse(is.na(LATE_AIRCRAFT_DELAY), 0, LATE_AIRCRAFT_DELAY),
    SECURITY_DELAY = ifelse(is.na(SECURITY_DELAY), 0, SECURITY_DELAY)
  ) %>%
  # Created a delayed flag and delayed means more than 15 minutes
  mutate(IS_DELAYED = ifelse(DEPARTURE_DELAY > 15, 1, 0))

# Here shows the structure of cleaned flights data
glimpse(flights_clean)
```

## Weather

```{r weather_cleaning}
# Selected only the columns we need for analysis
storm_clean <- storm_data %>%
  select(STATE, MONTH_NAME, EVENT_TYPE,
         DAMAGE_PROPERTY, DAMAGE_CROPS,
         INJURIES_DIRECT, DEATHS_DIRECT,
         BEGIN_DATE_TIME) %>%
  # Removed rows with no state
  filter(!is.na(STATE)) %>%
  # Standardized state to uppercase to match flights data
  mutate(STATE = toupper(STATE)) %>%
  # Replaced NA damage values with median 
  # Median is better than mean here because damage values
  # are heavily skewed by large disaster events
  mutate(
    DAMAGE_PROPERTY = ifelse(is.na(DAMAGE_PROPERTY),
                             median(DAMAGE_PROPERTY, na.rm = TRUE),
                             DAMAGE_PROPERTY),
    DAMAGE_CROPS = ifelse(is.na(DAMAGE_CROPS),
                          median(DAMAGE_CROPS, na.rm = TRUE),
                          DAMAGE_CROPS)
  )

# Shows structure of cleaned weather data
glimpse(storm_clean)
```

## Tweets

```{r tweets_cleaning}
# Convert to tibble for tidytext processing (Chapter 14)
twitter_tibble <- tibble(twitter_data)

# Tokenize words from tweet text
# unnest_tokens splits each tweet into individual words
twitter_unnested <- twitter_tibble %>%
  unnest_tokens(word, text)

# Custom stop words list these words appear in the Bing
# lexicon but dont carry meaningful sentiment in airline context
# For example "like" and "work" are too generic to be useful
# This was identified and added by Derek during analysis
stop_words_custom <- c("like", "work", "right", "well",
                       "enough", "free", "fine", "tough",
                       "afford", "cheaper", "rapid", "leverage")
stopwords_tibble <- tibble(word = stop_words_custom)

# Join with Bing sentiment lexicon (Chapter 15)
# Then remove custom stop words using anti_join
twitter_sentiment <- twitter_unnested %>%
  inner_join(get_sentiments("bing")) %>%
  anti_join(stopwords_tibble, by = "word") %>%
  count(word, sentiment, sort = TRUE)

# Separate positive and negative words for analysis
negative <- twitter_sentiment %>%
  filter(sentiment == "negative")

positive <- twitter_sentiment %>%
  filter(sentiment == "positive")

# Show top sentiment words
head(twitter_sentiment, 10)
```

# 4. Analysis

## Flights Patterns

```{r flights_by_airline}
# Which airlines have the most delays?
# Grouped summary using tidyverse (Chapter 12)
delays_by_airline <- flights_clean %>%
  group_by(AIRLINE_NAME) %>%
  summarise(
    total_flights = n(),
    total_delayed = sum(IS_DELAYED, na.rm = TRUE),
    delay_rate = round(total_delayed / total_flights * 100, 1),
    avg_delay_mins = round(mean(DEPARTURE_DELAY[DEPARTURE_DELAY > 0],
                                na.rm = TRUE), 1)
  ) %>%
  arrange(desc(delay_rate))

# Display as readable table
kable(delays_by_airline,
      col.names = c("Airline", "Total Flights",
                    "Delayed Flights", "Delay Rate %",
                    "Avg Delay (mins)"))
```

```{r flights_by_month}
# Which months have the most delays?
delays_by_month <- flights_clean %>%
  group_by(MONTH) %>%
  summarise(
    total_flights = n(),
    total_delayed = sum(IS_DELAYED, na.rm = TRUE),
    delay_rate = round(total_delayed / total_flights * 100, 1)
  ) %>%
  arrange(MONTH)

kable(delays_by_month,
      col.names = c("Month", "Total Flights",
                    "Delayed Flights", "Delay Rate %"))
```

## Weather Impact

```{r weather_by_state}
# Which states had the most storm events?
storms_by_state <- storm_clean %>%
  count(STATE, sort = TRUE) %>%
  head(15)

kable(storms_by_state,
      col.names = c("State", "Number of Storm Events"))
```

```{r weather_by_type}
# What types of storms occurred most often?
storms_by_type <- storm_clean %>%
  count(EVENT_TYPE, sort = TRUE) %>%
  head(10)

kable(storms_by_type,
      col.names = c("Storm Type", "Count"))
```

## Sentiment Analysis

```{r sentiment_summary}
# Count total positive vs negative words
sentiment_totals <- twitter_sentiment %>%
  group_by(sentiment) %>%
  summarise(total_words = sum(n))

kable(sentiment_totals,
      col.names = c("Sentiment", "Total Word Count"))
```

```{r negative_reasons}
# What are people most negative about?
# Uses the negativereason column from tweets dataset
negative_reasons <- twitter_data %>%
  filter(!is.na(negativereason) & negativereason != "") %>%
  count(negativereason, sort = TRUE)

kable(negative_reasons,
      col.names = c("Reason for Negative Tweet", "Count"))
```

# 5. Visualizations

## Flight Delay Rate by Airline

```{r graph_airline_delays}
# Bar chart showing delay rate per airline (Chapter 12 ggplot2)
# reorder() sorts bars by delay rate for easier reading
delays_by_airline %>%
  ggplot(aes(x = reorder(AIRLINE_NAME, delay_rate),
             y = delay_rate,
             fill = delay_rate)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +
  labs(
    title = "Flight Delay Rate by Airline (2015)",
    x = "Airline",
    y = "Delay Rate (%)",
    fill = "Delay Rate"
  ) +
  theme_minimal()
```

## Flight Delays by Month

```{r graph_monthly_delays}
# Line chart showing delay trends across months (Chapter 12)
# Helps identify seasonal patterns in delays
delays_by_month %>%
  ggplot(aes(x = MONTH, y = delay_rate)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "firebrick", size = 3) +
  scale_x_continuous(breaks = 1:12,
                     labels = month.abb) +
  labs(
    title = "Monthly Flight Delay Rate (2015)",
    x = "Month",
    y = "Delay Rate (%)"
  ) +
  theme_minimal()
```

## Top Storm States

```{r graph_storm_states}
# Bar chart of top 15 states by storm events (Chapter 12)
storms_by_state %>%
  ggplot(aes(x = reorder(STATE, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Top 15 States by Storm Events (2015)",
    x = "State",
    y = "Number of Storm Events",
    fill = "Events"
  ) +
  theme_minimal()
```

## Top Storm Types

```{r graph_storm_types}
# Bar chart of most common storm types
storms_by_type %>%
  ggplot(aes(x = reorder(EVENT_TYPE, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Most Common Storm Types (2015)",
    x = "Storm Type",
    y = "Count",
    fill = "Count"
  ) +
  theme_minimal()
```

## Tweet Sentiment - Top Positive Words

```{r graph_positive_words}
# Bar chart of top 15 positive words (Chapter 15)
positive %>%
  head(15) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 15 Positive Words in Airline Tweets",
    x = "Word",
    y = "Count"
  ) +
  theme_minimal()
```

## Tweet Sentiment - Top Negative Words

```{r graph_negative_words}
# Bar chart of top 15 negative words (Chapter 15)
negative %>%
  head(15) %>%
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Top 15 Negative Words in Airline Tweets",
    x = "Word",
    y = "Count"
  ) +
  theme_minimal()
```

## Why Are People Tweeting Negatively?

```{r graph_negative_reasons}
# Bar chart of negative tweet reasons
# This connects tweets directly to flight delay causes
negative_reasons %>%
  ggplot(aes(x = reorder(negativereason, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightsalmon", high = "firebrick") +
  labs(
    title = "Reasons for Negative Airline Tweets",
    x = "Reason",
    y = "Count",
    fill = "Count"
  ) +
  theme_minimal()
```

# 6. Integration

```{r integration}
# Here we connect weather events to flight delays by state

# Step 1: Get delay rate per state from flights data
state_delays <- flights_clean %>%
  filter(!is.na(ORIGIN_STATE)) %>%
  group_by(ORIGIN_STATE) %>%
  summarise(
    delay_rate = round(sum(IS_DELAYED) / n() * 100, 1),
    avg_weather_delay = round(mean(WEATHER_DELAY, na.rm = TRUE), 1)
  ) %>%
  # Convert state abbreviations to full uppercase names
  # to match the weather dataset format
  mutate(STATE_FULL = toupper(state.name[match(ORIGIN_STATE, 
                                                state.abb)]))

# Step 2: Get storm count per state from weather data
state_storms <- storm_clean %>%
  count(STATE) %>%
  rename(storm_count = n)

# Step 3: Join using full state names
weather_flight_join <- state_delays %>%
  filter(!is.na(STATE_FULL)) %>%
  inner_join(state_storms,
             by = c("STATE_FULL" = "STATE")) %>%
  arrange(desc(storm_count))

# Show results table
kable(head(weather_flight_join, 15),
      col.names = c("State Abbr", "Delay Rate %",
                    "Avg Weather Delay (mins)",
                    "State Full Name",
                    "Storm Events"))
```

```{r graph_integration}
# Scatter plot showing relationship between storms and delays
# Each point is a state and size shows average weather delay
weather_flight_join %>%
  ggplot(aes(x = storm_count,
             y = delay_rate,
             label = ORIGIN_STATE)) +
  geom_point(aes(size = avg_weather_delay),
             color = "steelblue", alpha = 0.7) +
  geom_text(vjust = -0.8, size = 3) +
  labs(
    title = "Storm Events vs Flight Delay Rate by State (2015)",
    x = "Number of Storm Events",
    y = "Flight Delay Rate (%)",
    size = "Avg Weather\nDelay (mins)"
  ) +
  theme_minimal()
```

# 7. Conclusion

**Goal of the mining work:**
This project aimed to uncover the relationship between weather
events, flight delays, and public sentiment about airlines in 2015.

**Patterns and insights found:**
- Certain airlines consistently had higher delay rates than others
- Delays peaked during specific months tied to weather seasons
- States with more storm events showed higher flight delay rates
- Public sentiment on Twitter was predominantly negative with
late flights and customer service being the top complaints
which directly mirrors the delay patterns found in flights data

**Why this matters:**
Understanding when and where delays happen and what causes
them helps airlines allocate resources better and helps
passengers make smarter travel decisions.

**Real world application:**
Airlines can use this analysis to proactively notify passengers
of likely delays based on weather forecasts and to identify
which routes need operational improvements.